---
title: "Pipeline A: 1% Quantile MNAR + k-NN MAR Imputation"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

## Overview

This notebook imputes missing values in VSN-normalized phosphoproteomics data
using a **1% quantile method for MNAR** values and **k-NN (k=10) for MAR**
values. Missing values are classified per genotype x timepoint group
(3 replicates each):

- **All 3 replicates missing -> MNAR**: imputed using each sample's 1st
  percentile intensity value (fixed low-value fill).
- **1-2 of 3 replicates missing -> MAR**: imputed using k-Nearest Neighbors
  (k = 10), borrowing values from similar peptides.

## 1. Install and Load Required Packages

```{r install-packages}
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")

required_bioc <- c("impute")
missing_bioc  <- required_bioc[!sapply(required_bioc, requireNamespace, quietly = TRUE)]
if (length(missing_bioc) > 0) BiocManager::install(missing_bioc, update = FALSE, ask = FALSE)

library(impute)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)

library(here)
library(openxlsx2)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(scales)
library(vsn)
library(impute)
library(writexl)

here::i_am("200_vsn_mnar_mar/200_vsn_mar_mnar_imputation_quantile.Rmd")
message("Loaded: ", nrow(data_raw), " rows x ", ncol(data_raw), " columns")
```



2. Identify sample columns

```{r identify-samples}
# FragPipe metadata columns to exclude
metadata_cols <- c(
  "Unique_site_ID", "Index", "Gene", "Protein", "Protein ID", "Peptide",
  "Best Localization Probability", "Final_protein_sites", "Residue", "Position", "n_sites_merged"
)

sample_cols <- setdiff(colnames(data_raw), metadata_cols)

# Verify numeric
data_test   <- data_raw %>% select(all_of(sample_cols))
sample_cols <- names(data_test)[sapply(data_test, function(col) {
  test_num <- suppressWarnings(as.numeric(as.character(col)))
  sum(!is.na(test_num)) / length(test_num) > 0.1
})]

message("Detected ", length(sample_cols), " sample columns")
message("First few: ", paste(head(sample_cols, 6), collapse = ", "))
```


## 3. Extract intensity matrix

```{r extract-matrix}
raw_mat <- data_raw %>%
  select(all_of(sample_cols)) %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(across(everything(), ~ifelse(. == 0 | is.na(.), NA, .))) %>%
  as.matrix()

rownames(raw_mat) <- data_raw$Unique_site_ID

message("Matrix: ", nrow(raw_mat), " sites x ", ncol(raw_mat), " samples")
message("Missing: ", sum(is.na(raw_mat)),
        " (", round(100 * sum(is.na(raw_mat)) / length(raw_mat), 1), "%)")
```

## 4. VSN normalization

```{r vsn-normalization}
vsn_mat <- vsn::justvsn(raw_mat)
message("VSN normalization complete")

if (!dir.exists(here("200_vsn_mnar_mar", "processed"))) {
  dir.create(here("200_vsn_mnar_mar", "processed"), recursive = TRUE)
}

png(here("200_vsn_mnar_mar", "processed", "200_vsn_meansd_plot.png"),
    width = 800, height = 600)
vsn::meanSdPlot(vsn_mat, main = "Mean-SD After VSN")
dev.off()
message("Saved: 200_vsn_meansd_plot.png")
```

## 4b. Save VSN-normalized non-imputed data

```{r save-vsn-nonimputed}
vsn_nonimputed_df <- as.data.frame(vsn_mat) %>%
  tibble::rownames_to_column("Unique_site_ID") %>%
  left_join(
    data_raw %>% select(Unique_site_ID, Index, Gene, Protein, `Protein ID`,
                        Peptide, `Best Localization Probability`,
                        Final_protein_sites),
    by = "Unique_site_ID"
  )

write_xlsx(vsn_nonimputed_df,
           here("200_vsn_mnar_mar", "processed",
                "200_vsn_normalized_nonimputed.xlsx"))
message("Saved: 200_vsn_normalized_nonimputed.xlsx")
```



## 3. Classify Missing Values (Cell-Level MNAR vs MAR)

```{r classify-missing}
# guy11/delMST7 design: 2 genotypes x 6 timepoints (0, 1h, 1_5h, 2h, 4h, 6h) x 3 reps = 36 samples
# e.g. guy11_0h_1, guy11_1_5h_2, mst7_6h_3
# Group = genotype_timepoint (everything except trailing _[rep])
conds <- gsub("_[0-9]+$", "", colnames(vsn_mat))

message("Unique condition groups: ", paste(unique(conds), collapse = ", "))

cat("Replicates per group:\n")
print(table(conds))

if (any(table(conds) != 3)) {
  warning("Some groups do not have exactly 3 replicates: ",
          paste(names(table(conds))[table(conds) != 3], collapse = ", "))
}

miss_mat <- is.na(vsn_mat)

MNAR_A <- matrix(FALSE, nrow(vsn_mat), ncol(vsn_mat), dimnames = dimnames(vsn_mat))

for (g in unique(conds)) {
  cc    <- which(conds == g)
  allna <- rowSums(miss_mat[, cc, drop = FALSE]) == length(cc)
  MNAR_A[allna, cc] <- TRUE
}

MNAR_A[!miss_mat] <- FALSE
MAR_A <- miss_mat & !MNAR_A

message("MNAR cells: ", sum(MNAR_A), " (", round(100*sum(MNAR_A)/sum(miss_mat), 1), "% of all missing)")
message("MAR cells:  ", sum(MAR_A),  " (", round(100*sum(MAR_A)/sum(miss_mat),  1), "% of all missing)")
```
 
## 4. Impute MNAR Values (1% Quantile)

```{r impute-mnar}
vsn_imp_quantile <- vsn_mat

q_low <- apply(vsn_mat, 2, function(x) quantile(x, probs = 0.01, na.rm = TRUE))
message("1% quantiles per sample:")
print(summary(q_low))

mnar_imputed_count <- 0
for (j in seq_len(ncol(vsn_imp_quantile))) {
  idx_mnar <- which(MNAR_A[, j])
  if (length(idx_mnar) > 0) {
    vsn_imp_quantile[idx_mnar, j] <- q_low[j]
    mnar_imputed_count <- mnar_imputed_count + length(idx_mnar)
  }
}

message("MNAR imputation complete: ", mnar_imputed_count, " values imputed")
```

## 5. Impute MAR Values (k-NN)

```{r impute-mar}
mar_missing_count <- sum(MAR_A & is.na(vsn_imp_quantile))

if (mar_missing_count > 0) {
  vsn_imp_quantile[MAR_A] <- NA
  vsn_mar_knn <- impute::impute.knn(vsn_imp_quantile, k = 10, rowmax = 0.9, colmax = 0.9)
  vsn_imp_quantile[MAR_A] <- vsn_mar_knn$data[MAR_A]
  message("MAR imputation complete: ", mar_missing_count, " values imputed (kNN k=10)")
} else {
  message("No MAR values to impute")
}

message("PIPELINE A final missing values: ", sum(is.na(vsn_imp_quantile)))
```


## 6. Summary Table

```{r summary-table}
Count <- c(sum(MNAR_A), sum(MAR_A), sum(MNAR_A) + sum(MAR_A), sum(is.na(vsn_imp_quantile)))

summary_tbl_A <- data.frame(
  Category     = c("MNAR (1% quantile)", "MAR (k-NN)", "Total imputed", "Remaining NA"),
  Count        = Count,
  Pct_of_total = round(100 * Count / (nrow(vsn_mat) * ncol(vsn_mat)), 2)
)

print(summary_tbl_A)
```



## 7. Save Imputed Matrix

```{r save-output}
dir.create(here::here("000_data_processing", "processed"), recursive = TRUE, showWarnings = FALSE)

saveRDS(vsn_imp_quantile, here::here("000_data_processing", "processed", "phospho_imputed_quantile.rds"))
write.csv(vsn_imp_quantile, here::here("000_data_processing", "processed", "phospho_imputed_quantile.csv"))

message("Saved Pipeline A imputed matrix to 000_data_processing/processed/phospho_imputed_quantile.rds and .csv")
```


8. Prepare and Save Output
```{r prepare-output}
final_df <- as.data.frame(vsn_imp_quantile) %>%
  tibble::rownames_to_column("Unique_site_ID") %>%
  left_join(
    data_raw %>% select(Unique_site_ID, Index, Gene, Protein, `Protein ID`,
                        Peptide, `Best Localization Probability`,
                        Final_protein_sites),
    by = "Unique_site_ID"
  )

message("Final dataset: ", nrow(final_df), " rows x ", ncol(final_df), " columns")
```

9. Save Wide Format

```{r save-wide}
write_xlsx(final_df,
           here("200_vsn_mnar_mar", "processed", "200_vsn_mar_mnar_imputed.xlsx"))
message("Saved: 200_vsn_mar_mnar_imputed.xlsx")
```

10. Create and Save Long Format

```{r create-long}
final_df_long <- final_df %>%
  select(Unique_site_ID, `Protein ID`, Gene, Peptide,
         Final_protein_sites, all_of(sample_cols)) %>%
  pivot_longer(
    cols      = all_of(sample_cols),
    names_to  = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(
    # guy11/mst7 format: guy11_0h_1, guy11_1_5h_2, mst7_6h_3
    Condition = case_when(
      str_detect(Sample, "^guy11_") ~ "guy11",
      str_detect(Sample, "^mst7_")  ~ "mst7",
      TRUE ~ str_extract(Sample, "^[^_]+")
    ),
    Time = str_extract(Sample, "(?<=_)[0-9]+(_[0-9]+)?h(?=_)") %>%
      str_remove("h$") %>%
      str_replace("_", ".") %>%
      as.numeric(),
    Rep  = str_extract(Sample, "[0-9]+$") %>% as.numeric()
  ) %>%
  select(Unique_site_ID, Protein = `Protein ID`, Gene, Peptide,
         Sample, Condition, Time, Rep, Intensity, Final_protein_sites)

message("Long format: ", nrow(final_df_long), " rows")

write_xlsx(final_df_long,
           here("200_vsn_mnar_mar", "processed",
                "200_vsn_mar_mnar_imputed_long.xlsx"))
message("Saved: 200_vsn_mar_mnar_imputed_long.xlsx")
```



11 Visualisation: Normalisation Effect

```{r visualize-normalization}
condition_colors <- c(
  "guy11" = "#4E79A7",
  "mst7"  = "#F28E2B"
)

before_totals <- colSums(raw_mat, na.rm = TRUE) %>%
  enframe(name = "Sample", value = "Total_Intensity") %>%
  mutate(Stage     = "1. Raw Data",
         Condition = case_when(
           str_detect(Sample, "^guy11_") ~ "guy11",
           str_detect(Sample, "^mst7_")  ~ "mst7",
           TRUE ~ "guy11"
         ))

after_totals <- colSums(2^vsn_imp_quantile, na.rm = TRUE) %>%
  enframe(name = "Sample", value = "Total_Intensity") %>%
  mutate(Stage     = "2. After VSN + Imputation",
         Condition = case_when(
           str_detect(Sample, "^guy11_") ~ "guy11",
           str_detect(Sample, "^mst7_")  ~ "mst7",
           TRUE ~ "guy11"
         ))

p_norm <- bind_rows(before_totals, after_totals) %>%
  group_by(Stage) %>%
  mutate(mean_intensity = mean(Total_Intensity, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = Sample, y = Total_Intensity, fill = Condition)) +
  geom_col() +
  geom_hline(aes(yintercept = mean_intensity),
             linetype = "dashed", color = "black", linewidth = 0.8) +
  scale_fill_manual(values = condition_colors) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~Stage, ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7)) +
  labs(title    = "Total Intensity: VSN Normalisation Effect",
       subtitle = "Dashed line = mean total intensity",
       y = "Total Intensity", x = "Sample")

print(p_norm)
ggsave(here("200_vsn_mnar_mar", "processed", "200_vsn_normalization_effect.png"),
       p_norm, width = 12, height = 8, dpi = 300)
message("Saved: 200_vsn_normalization_effect.png")
```






```{r visualize-imputation}
dist_combined <- bind_rows(
  data.frame(Intensity = as.vector(vsn_mat), Stage = "Before Imputation") %>%
    filter(!is.na(Intensity)),
  data.frame(Intensity = as.vector(vsn_imp_quantile), Stage = "After Imputation")
) %>%
  mutate(Stage = factor(Stage, levels = c("Before Imputation", "After Imputation")))

p_imp <- ggplot(dist_combined, aes(x = Intensity, fill = Stage)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  facet_wrap(~Stage, ncol = 1) +
  scale_fill_manual(values = c("#0d47a1", "#b71c1c")) +
  theme_minimal(base_size = 14) +
  labs(title    = "Intensity Distribution: Imputation Effect",
       subtitle = "MAR: kNN; MNAR: 1% quantile",
       x = "Intensity (VSN scale)", y = "Frequency") +
  theme(legend.position = "none")

print(p_imp)
ggsave(here("200_vsn_mnar_mar", "processed", "200_imputation_effect.png"),
       p_imp, width = 10, height = 8, dpi = 300)
message("Saved: 200_imputation_effect.png")
```






```{r final-summary}
mnar_rows <- rowSums(MNAR_A) > 0

cat("\n===============================================\n")
cat("VSN NORMALISATION & IMPUTATION SUMMARY\n")
cat("Appressorium guy11/mst7 Time Course\n")
cat("===============================================\n\n")
cat(sprintf("Sites: %d | Samples: %d\n", nrow(raw_mat), ncol(raw_mat)))
cat(sprintf("Missing before: %.1f%%\n\n",
            100 * sum(is.na(raw_mat)) / length(raw_mat)))
cat(sprintf("MNAR values: %d (%.1f%% of missing) -> imputed with 1%% quantile\n",
            sum(MNAR_A), 100*sum(MNAR_A)/sum(miss_mat)))
cat(sprintf("MAR values:  %d (%.1f%% of missing) -> imputed with kNN (k=10)\n",
            sum(MAR_A), 100*sum(MAR_A)/sum(miss_mat)))
cat(sprintf("Sites with >=1 fully-missing genotype x timepoint block: %d (%.1f%%)\n\n",
            sum(mnar_rows), 100*sum(mnar_rows)/nrow(miss_mat)))
cat("Output files:\n")
cat("  200_vsn_normalized_nonimputed.xlsx\n")
cat("  200_vsn_mar_mnar_imputed.xlsx\n")
cat("  200_vsn_mar_mnar_imputed_long.xlsx\n")
cat("===============================================\n")
```






















## Session Info

```{r session-info}
sessionInfo()
```



