---
title: "Infection phosphoproteome analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Source helper scripts

```{r source-scripts}
source("analysis/scripts/phosphoproteome_cleanup.r")
source("analysis/scripts/phosphoproteome_analysis.r")
```

## File paths

```{r paths}
pmk1_path <- file.path("raw", "pmk1", "combined_modified_peptide.tsv")
mst7_path <- file.path("raw", "mst7", "combined_modified_peptide.tsv")

# Helper to detect spectral and MaxLFQ columns from a data frame
detect_spec_and_lfq <- function(df) {
  spec_cols <- colnames(df)[stringr::str_detect(colnames(df), regex("Spectral Count", ignore_case = TRUE))]
  lfq_cols <- colnames(df)[stringr::str_detect(colnames(df), regex("MaxLFQ Intensity", ignore_case = TRUE))]
  list(spec = spec_cols, lfq = lfq_cols)
}
```

## Cleanup datasets

```{r cleanup}
# Load raw to detect columns
raw_pmk1 <- load_phospho_data(pmk1_path)
cols_pmk1 <- detect_spec_and_lfq(raw_pmk1)

raw_mst7 <- load_phospho_data(mst7_path)
cols_mst7 <- detect_spec_and_lfq(raw_mst7)

# Run full cleanup (phospho_all and phospho_only)
pmk1_clean <- cleanup_phospho_dataset(pmk1_path, mod_col = "Assigned Modifications")
mst7_clean <- cleanup_phospho_dataset(mst7_path, mod_col = "Assigned Modifications")

# Expose datasets
pmk1_phospho_all <- pmk1_clean$phospho_all
pmk1_phospho_only <- pmk1_clean$phospho_only

mst7_phospho_all <- mst7_clean$phospho_all
mst7_phospho_only <- mst7_clean$phospho_only
```

## Summaries and plots (phospho_all)

```{r summaries-phospho-all}
# Summarise per-sample metrics for phospho_all
pmk1_summary_all <- summarise_phospho_samples(pmk1_phospho_all, spec_cols = cols_pmk1$spec, lfq_cols = cols_pmk1$lfq)
mst7_summary_all <- summarise_phospho_samples(mst7_phospho_all, spec_cols = cols_mst7$spec, lfq_cols = cols_mst7$lfq)

# Example plots
plot_total_spectral_counts(pmk1_summary_all)
plot_total_maxlfq(pmk1_summary_all)
plot_total_phospho_peptides(pmk1_summary_all)
plot_total_phospho_proteins(pmk1_summary_all)

plot_total_spectral_counts(mst7_summary_all)
plot_total_maxlfq(mst7_summary_all)
plot_total_phospho_peptides(mst7_summary_all)
plot_total_phospho_proteins(mst7_summary_all)
```

```{r save-summaries-phospho-all}
# Create output folder if it doesn't exist
if (!dir.exists("results")) dir.create("results")

## ---- Save summaries for phospho_all ----

write.csv(pmk1_summary_all, "results/pmk1_summary_all.csv", row.names = FALSE)
write.csv(mst7_summary_all, "results/mst7_summary_all.csv", row.names = FALSE)

# PMK1 plots
ggplot2::ggsave("results/pmk1_total_spectral_counts_all.png",
                plot_total_spectral_counts(pmk1_summary_all), width = 7, height = 5)

ggplot2::ggsave("results/pmk1_total_maxlfq_all.png",
                plot_total_maxlfq(pmk1_summary_all), width = 7, height = 5)

ggplot2::ggsave("results/pmk1_total_phospho_peptides_all.png",
                plot_total_phospho_peptides(pmk1_summary_all), width = 7, height = 5)

ggplot2::ggsave("results/pmk1_total_phospho_proteins_all.png",
                plot_total_phospho_proteins(pmk1_summary_all), width = 7, height = 5)

# MST7 plots
ggplot2::ggsave("results/mst7_total_spectral_counts_all.png",
                plot_total_spectral_counts(mst7_summary_all), width = 7, height = 5)

ggplot2::ggsave("results/mst7_total_maxlfq_all.png",
                plot_total_maxlfq(mst7_summary_all), width = 7, height = 5)

ggplot2::ggsave("results/mst7_total_phospho_peptides_all.png",
                plot_total_phospho_peptides(mst7_summary_all), width = 7, height = 5)

ggplot2::ggsave("results/mst7_total_phospho_proteins_all.png",
                plot_total_phospho_proteins(mst7_summary_all), width = 7, height = 5)
```

## Summaries and plots (phospho_only)

```{r summaries-phospho-only}
# Summarise per-sample metrics for phospho_only
pmk1_summary_only <- summarise_phospho_samples(pmk1_phospho_only, spec_cols = cols_pmk1$spec, lfq_cols = cols_pmk1$lfq)
mst7_summary_only <- summarise_phospho_samples(mst7_phospho_only, spec_cols = cols_mst7$spec, lfq_cols = cols_mst7$lfq)

# Example plots
plot_total_spectral_counts(pmk1_summary_only)
plot_total_maxlfq(pmk1_summary_only)
plot_total_phospho_peptides(pmk1_summary_only)
plot_total_phospho_proteins(pmk1_summary_only)

plot_total_spectral_counts(mst7_summary_only)
plot_total_maxlfq(mst7_summary_only)
plot_total_phospho_peptides(mst7_summary_only)
plot_total_phospho_proteins(mst7_summary_only)
```

```{r save-summaries-phospho-only}
# Create output folder if it doesn't exist
if (!dir.exists("results")) dir.create("results")

## ---- Save summaries for phospho_only ----

write.csv(pmk1_summary_only, "results/pmk1_summary_only.csv", row.names = FALSE)
write.csv(mst7_summary_only, "results/mst7_summary_only.csv", row.names = FALSE)

# PMK1 plots
ggplot2::ggsave("results/pmk1_total_spectral_counts_only.png",
                plot_total_spectral_counts(pmk1_summary_only), width = 7, height = 5)

ggplot2::ggsave("results/pmk1_total_maxlfq_only.png",
                plot_total_maxlfq(pmk1_summary_only), width = 7, height = 5)

ggplot2::ggsave("results/pmk1_total_phospho_peptides_only.png",
                plot_total_phospho_peptides(pmk1_summary_only), width = 7, height = 5)

ggplot2::ggsave("results/pmk1_total_phospho_proteins_only.png",
                plot_total_phospho_proteins(pmk1_summary_only), width = 7, height = 5)

# MST7 plots
ggplot2::ggsave("results/mst7_total_spectral_counts_only.png",
                plot_total_spectral_counts(mst7_summary_only), width = 7, height = 5)

ggplot2::ggsave("results/mst7_total_maxlfq_only.png",
                plot_total_maxlfq(mst7_summary_only), width = 7, height = 5)

ggplot2::ggsave("results/mst7_total_phospho_peptides_only.png",
                plot_total_phospho_peptides(mst7_summary_only), width = 7, height = 5)

ggplot2::ggsave("results/mst7_total_phospho_proteins_only.png",
                plot_total_phospho_proteins(mst7_summary_only), width = 7, height = 5)
```

## Save cleaned data frames

```{r save-cleaned-dfs}
# Create output folder if it doesn't exist
if (!dir.exists("results")) dir.create("results")

# PMK1
write.csv(pmk1_phospho_all,  "results/pmk1_phospho_all.csv",  row.names = FALSE)
write.csv(pmk1_phospho_only, "results/pmk1_phospho_only.csv", row.names = FALSE)

# MST7
write.csv(mst7_phospho_all,  "results/mst7_phospho_all.csv",  row.names = FALSE)
write.csv(mst7_phospho_only, "results/mst7_phospho_only.csv", row.names = FALSE)
```

## Venn analyses

```{r venn-sets}
# A feature is detected when at least one sample for that strain has a count > 0.
features_for_strain <- function(df, spec_cols, strain, id_col) {
  strain_cols <- spec_cols[stringr::str_detect(
    tolower(spec_cols), paste0("^", stringr::fixed(tolower(strain)), "_")
  )]
  if (length(strain_cols) == 0) stop("No spectral-count columns found for ", strain)

  df %>%
    filter(rowSums(across(all_of(strain_cols), ~ replace_na(as.numeric(.x), 0))) > 0) %>%
    pull(all_of(id_col)) %>%
    as.character() %>%
    na.omit() %>%
    unique()
}

make_comparison <- function(df_a, spec_a, strain_a, df_b, spec_b, strain_b,
                            label_a = strain_a, label_b = strain_b) {
  setNames(list(
    peptides = setNames(list(
      features_for_strain(df_a, spec_a, strain_a, "Peptide Sequence"),
      features_for_strain(df_b, spec_b, strain_b, "Peptide Sequence")
    ), c(label_a, label_b)),
    proteins = setNames(list(
      features_for_strain(df_a, spec_a, strain_a, "Protein ID"),
      features_for_strain(df_b, spec_b, strain_b, "Protein ID")
    ), c(label_a, label_b))
  ), c("peptides", "proteins"))
}

# Within-experiment comparisons.
venn_pmk1 <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "guy11",
                             pmk1_phospho_all, cols_pmk1$spec, "pmk1")
venn_mst7 <- make_comparison(mst7_phospho_all, cols_mst7$spec, "guy11",
                             mst7_phospho_all, cols_mst7$spec, "mst7")

# Between-experiment comparisons, using all samples for each named strain.
venn_guy11 <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "guy11",
                              mst7_phospho_all, cols_mst7$spec, "guy11",
                              "Guy11 PMK1", "Guy11 MST7")
venn_mutants <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "pmk1",
                                mst7_phospho_all, cols_mst7$spec, "mst7")
```
```

## Plot proportional Venn diagrams

```{r plot-venns}
library(VennDiagram)

plot_comparison <- function(comparison, entity, title) {
  sets <- comparison[[entity]]
  counts <- c(length(sets[[1]]), length(sets[[2]]),
              length(intersect(sets[[1]], sets[[2]])))
  VennDiagram::draw.pairwise.venn(
    area1 = counts[[1]], area2 = counts[[2]], cross.area = counts[[3]],
    category = names(sets), scaled = TRUE, euler.d = TRUE,
    fill = c("#2F6690", "#F28E2B"), alpha = c(0.55, 0.55),
    cex = 1.1, cat.cex = 1.1, ind = FALSE
  )
  grid::grid.draw(grid::textGrob(
    title, y = grid::unit(0.98, "npc"),
    just = c("center", "top"),
    gp = grid::gpar(fontface = "bold", cex = 1.1)
  ))
  invisible(NULL)
}

plot_comparison(venn_pmk1, "peptides", "PMK1 experiment: peptides")
plot_comparison(venn_pmk1, "proteins", "PMK1 experiment: phosphoproteins")
plot_comparison(venn_mst7, "peptides", "MST7 experiment: peptides")
plot_comparison(venn_mst7, "proteins", "MST7 experiment: phosphoproteins")
plot_comparison(venn_guy11, "peptides", "Guy11 across experiments: peptides")
plot_comparison(venn_guy11, "proteins", "Guy11 across experiments: phosphoproteins")
plot_comparison(venn_mutants, "peptides", "PMK1 vs MST7: peptides")
plot_comparison(venn_mutants, "proteins", "PMK1 vs MST7: phosphoproteins")
```

## Save Venn diagrams as PNG files

```{r save-venns}
venn_plots <- list(
  pmk1_peptides = list(comparison = venn_pmk1, entity = "peptides",
                       title = "PMK1 experiment: peptides"),
  pmk1_phosphoproteins = list(comparison = venn_pmk1, entity = "proteins",
                              title = "PMK1 experiment: phosphoproteins"),
  mst7_peptides = list(comparison = venn_mst7, entity = "peptides",
                       title = "MST7 experiment: peptides"),
  mst7_phosphoproteins = list(comparison = venn_mst7, entity = "proteins",
                              title = "MST7 experiment: phosphoproteins"),
  guy11_peptides = list(comparison = venn_guy11, entity = "peptides",
                        title = "Guy11 across experiments: peptides"),
  guy11_phosphoproteins = list(comparison = venn_guy11, entity = "proteins",
                               title = "Guy11 across experiments: phosphoproteins"),
  mutants_peptides = list(comparison = venn_mutants, entity = "peptides",
                          title = "PMK1 vs MST7: peptides"),
  mutants_phosphoproteins = list(comparison = venn_mutants, entity = "proteins",
                                title = "PMK1 vs MST7: phosphoproteins")
)

for (plot_name in names(venn_plots)) {
  plot_spec <- venn_plots[[plot_name]]
  grDevices::png(
    filename = file.path("results", paste0("venn_", plot_name, ".png")),
    width = 7, height = 6, units = "in", res = 300
  )
  plot_comparison(plot_spec$comparison, plot_spec$entity, plot_spec$title)
  grDevices::dev.off()
}
```


## Notes

- The functions assume the column names contain the strings `Spectral Count` and `MaxLFQ Intensity` to detect sample columns. Adjust detection if your TSV differs.

