---
title: "Infection phosphoproteome analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Source helper scripts

```{r source-scripts}
source("analysis/scripts/phosphoproteome_cleanup.r")
source("analysis/scripts/phosphoproteome_analysis.r")
source("analysis/scripts/phosphoproteome_vsn_imputation.r")
source("analysis/scripts/phosphoproteome_diffexp_bootstrap.r")
```

## File paths

```{r paths}
pmk1_path <- file.path("raw", "pmk1", "combined_modified_peptide.tsv")
mst7_path <- file.path("raw", "mst7", "combined_modified_peptide.tsv")

# Helper to detect spectral and MaxLFQ columns from a data frame
detect_spec_and_lfq <- function(df) {
  spec_cols <- colnames(df)[stringr::str_detect(colnames(df), regex("Spectral Count", ignore_case = TRUE))]
  lfq_cols <- colnames(df)[stringr::str_detect(colnames(df), regex("MaxLFQ Intensity", ignore_case = TRUE))]
  intensity_cols <- colnames(df)[
    stringr::str_detect(colnames(df), regex(" Intensity$", ignore_case = TRUE)) &
      !stringr::str_detect(colnames(df), regex(" MaxLFQ Intensity$", ignore_case = TRUE))
  ]
  list(spec = spec_cols, lfq = lfq_cols, intensity = intensity_cols)
}
```

## Cleanup datasets

```{r cleanup pmk1}
dir.create("results", showWarnings = FALSE, recursive = TRUE)

raw_pmk1 <- load_phospho_data(pmk1_path)
cols_pmk1 <- detect_spec_and_lfq(raw_pmk1)

pmk1_clean <- cleanup_phospho_dataset(pmk1_path, mod_col = "Assigned Modifications")


pmk1_phospho_all_unfiltered <- pmk1_clean$phospho_all
pmk1_phospho_only_unfiltered <- pmk1_clean$phospho_only
pmk1_phospho_all <- pmk1_clean$phospho_all_biorep_filtered





writexl::write_xlsx(raw_pmk1, "results/pmk1_00_raw.xlsx")
writexl::write_xlsx(pmk1_phospho_all_unfiltered, "results/pmk1_01_phospho_all.xlsx")
writexl::write_xlsx(pmk1_phospho_only_unfiltered, "results/pmk1_02_phospho_only.xlsx")
writexl::write_xlsx(pmk1_phospho_all, "results/pmk1_03_phospho_all_min_2_bioreps.xlsx")
```

```{r cleanup mst7}
dir.create("results", showWarnings = FALSE, recursive = TRUE)

raw_mst7 <- load_phospho_data(mst7_path)
cols_mst7 <- detect_spec_and_lfq(raw_mst7)
mst7_clean <- cleanup_phospho_dataset(mst7_path, mod_col = "Assigned Modifications")


mst7_phospho_all_unfiltered <- mst7_clean$phospho_all
mst7_phospho_only_unfiltered <- mst7_clean$phospho_only
mst7_phospho_all <- mst7_clean$phospho_all_biorep_filtered


writexl::write_xlsx(raw_mst7, "results/mst7_00_raw.xlsx")
writexl::write_xlsx(mst7_phospho_all_unfiltered, "results/mst7_01_phospho_all.xlsx")
writexl::write_xlsx(mst7_phospho_only_unfiltered, "results/mst7_02_phospho_only.xlsx")
writexl::write_xlsx(mst7_phospho_all, "results/mst7_03_phospho_all_min_2_bioreps.xlsx")

```

## VSN normalisation and imputation

```{r vsn-imputation-pmk1}
pmk1_vsn <- process_vsn_imputation(
  "results/pmk1_03_phospho_all_min_2_bioreps.xlsx", "pmk1"
)
```

```{r vsn-imputation-mst7}
mst7_vsn <- process_vsn_imputation(
  "results/mst7_03_phospho_all_min_2_bioreps.xlsx", "mst7"
)
```

## Reload saved processing stages

```{r reload-processing-stages}
pmk1_filtered <- readxl::read_xlsx("results/pmk1_03_phospho_all_min_2_bioreps.xlsx")
pmk1_vsn_normalized <- readxl::read_xlsx("results/pmk1_05_vsn_normalized_nonimputed.xlsx")
pmk1_imputed <- readxl::read_xlsx("results/pmk1_05_vsn_mar_mnar_imputed.xlsx")

mst7_filtered <- readxl::read_xlsx("results/mst7_03_phospho_all_min_2_bioreps.xlsx")
mst7_vsn_normalized <- readxl::read_xlsx("results/mst7_05_vsn_normalized_nonimputed.xlsx")
mst7_imputed <- readxl::read_xlsx("results/mst7_05_vsn_mar_mnar_imputed.xlsx")

pmk1_sample_cols <- stringr::str_remove(
  cols_pmk1$intensity, " Intensity$"
)
mst7_sample_cols <- stringr::str_remove(
  cols_mst7$intensity, " Intensity$"
)
```

## Summaries and plots (all processing stages)

```{r summaries-all-stages}
pmk1_summary_filtered <- summarise_phospho_stage(
  pmk1_filtered, "pmk1", "filtered",
  spec_cols = cols_pmk1$spec, lfq_cols = cols_pmk1$lfq,
  intensity_cols = cols_pmk1$intensity
)
pmk1_summary_normalized <- summarise_phospho_stage(
  pmk1_vsn_normalized, "pmk1", "normalized",
  intensity_cols = pmk1_sample_cols
)
pmk1_summary_imputed <- summarise_phospho_stage(
  pmk1_imputed, "pmk1", "imputed",
  intensity_cols = pmk1_sample_cols
)

mst7_summary_filtered <- summarise_phospho_stage(
  mst7_filtered, "mst7", "filtered",
  spec_cols = cols_mst7$spec, lfq_cols = cols_mst7$lfq,
  intensity_cols = cols_mst7$intensity
)
mst7_summary_normalized <- summarise_phospho_stage(
  mst7_vsn_normalized, "mst7", "normalized",
  intensity_cols = mst7_sample_cols
)
mst7_summary_imputed <- summarise_phospho_stage(
  mst7_imputed, "mst7", "imputed",
  intensity_cols = mst7_sample_cols
)

summary_all_stages <- dplyr::bind_rows(
  pmk1_summary_filtered, pmk1_summary_normalized, pmk1_summary_imputed,
  mst7_summary_filtered, mst7_summary_normalized, mst7_summary_imputed
)

writexl::write_xlsx(summary_all_stages,
                    "results/summary_phospho_all_stages.xlsx")
ggplot2::ggsave("results/summary_total_spectral_counts_by_stage.png",
                plot_stage_spectral_counts(summary_all_stages), width = 12, height = 7)
ggplot2::ggsave("results/summary_total_maxlfq_by_stage.png",
                plot_stage_maxlfq(summary_all_stages), width = 12, height = 7)
ggplot2::ggsave("results/summary_total_intensity_by_stage.png",
                plot_stage_intensity(summary_all_stages), width = 12, height = 7)
ggplot2::ggsave("results/summary_phospho_peptides_by_stage.png",
                plot_stage_phospho_peptides(summary_all_stages), width = 12, height = 7)
ggplot2::ggsave("results/summary_phospho_proteins_by_stage.png",
                plot_stage_phospho_proteins(summary_all_stages), width = 12, height = 7)
```

## Peptide intensity distribution by sample (before vs after VSN/imputation)

One boxplot per processing stage and experiment: x-axis is sample (same
sample order/layout as the other per-sample QC plots above), y-axis is
intensity, and each point is one peptide's intensity in that sample. The
filtered stage is on the raw intensity scale (zeros, i.e. "not observed",
are dropped, and the axis is log10 since raw intensities span several
orders of magnitude even within a single sample); the normalized/imputed
stages are on the VSN scale (already roughly log-transformed, so plotted
linearly), where all values -- including zero or negative -- are kept.

```{r intensity-boxplots-by-sample}
intensity_boxplot_specs <- list(
  list(experiment = "pmk1", stage = "filtered",
       df = pmk1_filtered, intensity_cols = cols_pmk1$intensity,
       intensity_suffix = " Intensity", drop_nonpositive = TRUE,
       y_label = "Intensity (log10 scale)"),
  list(experiment = "pmk1", stage = "normalized",
       df = pmk1_vsn_normalized, intensity_cols = pmk1_sample_cols,
       intensity_suffix = "", drop_nonpositive = FALSE,
       y_label = "VSN-normalized intensity"),
  list(experiment = "pmk1", stage = "imputed",
       df = pmk1_imputed, intensity_cols = pmk1_sample_cols,
       intensity_suffix = "", drop_nonpositive = FALSE,
       y_label = "VSN-normalized, imputed intensity"),
  list(experiment = "mst7", stage = "filtered",
       df = mst7_filtered, intensity_cols = cols_mst7$intensity,
       intensity_suffix = " Intensity", drop_nonpositive = TRUE,
       y_label = "Intensity (log10 scale)"),
  list(experiment = "mst7", stage = "normalized",
       df = mst7_vsn_normalized, intensity_cols = mst7_sample_cols,
       intensity_suffix = "", drop_nonpositive = FALSE,
       y_label = "VSN-normalized intensity"),
  list(experiment = "mst7", stage = "imputed",
       df = mst7_imputed, intensity_cols = mst7_sample_cols,
       intensity_suffix = "", drop_nonpositive = FALSE,
       y_label = "VSN-normalized, imputed intensity")
)

for (spec in intensity_boxplot_specs) {
  plot_title <- paste0(
    toupper(spec$experiment), " experiment (", spec$stage,
    "): peptide intensity by sample"
  )
  p <- plot_intensity_boxplot_by_sample(
    spec$df, spec$intensity_cols,
    intensity_suffix = spec$intensity_suffix,
    drop_nonpositive = spec$drop_nonpositive,
    title = plot_title, y_label = spec$y_label
  )
  print(p)
  ggplot2::ggsave(
    file.path("results", paste0(spec$experiment, "_", spec$stage,
                                "_intensity_boxplot_by_sample.png")),
    p, width = 14, height = 6
  )
}
```

## Phospho-only comparison summary

```{r phospho-only-summary}
pmk1_only <- readxl::read_xlsx(
  "results/pmk1_04_phospho_only_min_2_bioreps.xlsx"
)
mst7_only <- readxl::read_xlsx(
  "results/mst7_04_phospho_only_min_2_bioreps.xlsx"
)

pmk1_summary_only <- summarise_phospho_stage(
  pmk1_only, "pmk1", "phospho_only",
  spec_cols = cols_pmk1$spec, lfq_cols = cols_pmk1$lfq,
  intensity_cols = cols_pmk1$intensity
)
mst7_summary_only <- summarise_phospho_stage(
  mst7_only, "mst7", "phospho_only",
  spec_cols = cols_mst7$spec, lfq_cols = cols_mst7$lfq,
  intensity_cols = cols_mst7$intensity
)

summary_comparison <- dplyr::bind_rows(
  summary_all_stages, pmk1_summary_only, mst7_summary_only
)
overall_summary_comparison <- overall_phospho_summary(summary_comparison)

writexl::write_xlsx(pmk1_summary_only, "results/pmk1_summary_phospho_only.xlsx")
writexl::write_xlsx(mst7_summary_only, "results/mst7_summary_phospho_only.xlsx")
writexl::write_xlsx(overall_summary_comparison,
                    "results/overall_summary_comparison.xlsx")
```



## Venn analyses

```{r venn-sets}
# A feature is detected when at least one sample for that strain has a count > 0.
features_for_strain <- function(df, spec_cols, strain, id_col) {
  strain_cols <- spec_cols[stringr::str_detect(
    tolower(spec_cols), paste0("^", stringr::fixed(tolower(strain)), "_")
  )]
  if (length(strain_cols) == 0) stop("No spectral-count columns found for ", strain)

  df %>%
    filter(rowSums(across(all_of(strain_cols), ~ replace_na(as.numeric(.x), 0))) > 0) %>%
    pull(all_of(id_col)) %>%
    as.character() %>%
    na.omit() %>%
    unique()
}

make_comparison <- function(df_a, spec_a, strain_a, df_b, spec_b, strain_b,
                            label_a = strain_a, label_b = strain_b) {
  setNames(list(
    peptides = setNames(list(
      features_for_strain(df_a, spec_a, strain_a, "Peptide Sequence"),
      features_for_strain(df_b, spec_b, strain_b, "Peptide Sequence")
    ), c(label_a, label_b)),
    proteins = setNames(list(
      features_for_strain(df_a, spec_a, strain_a, "Protein ID"),
      features_for_strain(df_b, spec_b, strain_b, "Protein ID")
    ), c(label_a, label_b))
  ), c("peptides", "proteins"))
}

# Within-experiment comparisons.
venn_pmk1 <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "guy11",
                             pmk1_phospho_all, cols_pmk1$spec, "pmk1")
venn_mst7 <- make_comparison(mst7_phospho_all, cols_mst7$spec, "guy11",
                             mst7_phospho_all, cols_mst7$spec, "mst7")

# Between-experiment comparisons, using all samples for each named strain.
venn_guy11 <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "guy11",
                              mst7_phospho_all, cols_mst7$spec, "guy11",
                              "Guy11 PMK1", "Guy11 MST7")
venn_mutants <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "pmk1",
                                mst7_phospho_all, cols_mst7$spec, "mst7")

# Restrict the mutant comparison to features detected in both Guy11 datasets.
guy11_shared <- list(
  peptides = intersect(venn_guy11$peptides[[1]], venn_guy11$peptides[[2]]),
  proteins = intersect(venn_guy11$proteins[[1]], venn_guy11$proteins[[2]])
)

venn_mutants_guy11_shared <- make_comparison(
  mst7_phospho_all %>%
    filter(as.character(.data[["Peptide Sequence"]]) %in% guy11_shared$peptides),
  cols_mst7$spec, "mst7",
  pmk1_phospho_all %>%
    filter(as.character(.data[["Peptide Sequence"]]) %in% guy11_shared$peptides),
  cols_pmk1$spec, "pmk1", "MST7", "PMK1"
)

venn_mutants_guy11_shared$proteins <- setNames(list(
  features_for_strain(
    mst7_phospho_all %>%
      filter(as.character(.data[["Protein ID"]]) %in% guy11_shared$proteins),
    cols_mst7$spec, "mst7", "Protein ID"
  ),
  features_for_strain(
    pmk1_phospho_all %>%
      filter(as.character(.data[["Protein ID"]]) %in% guy11_shared$proteins),
    cols_pmk1$spec, "pmk1", "Protein ID"
  )
), c("MST7", "PMK1"))
```
```

## Plot proportional Venn diagrams

```{r plot-venns}
library(VennDiagram)

plot_comparison <- function(comparison, entity, title) {
  sets <- comparison[[entity]]
  counts <- c(length(sets[[1]]), length(sets[[2]]),
              length(intersect(sets[[1]], sets[[2]])))
  VennDiagram::draw.pairwise.venn(
    area1 = counts[[1]], area2 = counts[[2]], cross.area = counts[[3]],
    category = names(sets), scaled = TRUE, euler.d = TRUE,
    fill = c("#2F6690", "#F28E2B"), alpha = c(0.55, 0.55),
    cex = 1.1, cat.cex = 1.1, ind = FALSE
  )
  grid::grid.draw(grid::textGrob(
    title, y = grid::unit(0.98, "npc"),
    just = c("center", "top"),
    gp = grid::gpar(fontface = "bold", cex = 1.1)
  ))
  invisible(NULL)
}

plot_comparison(venn_pmk1, "peptides", "PMK1 experiment: peptides")
plot_comparison(venn_pmk1, "proteins", "PMK1 experiment: phosphoproteins")
plot_comparison(venn_mst7, "peptides", "MST7 experiment: peptides")
plot_comparison(venn_mst7, "proteins", "MST7 experiment: phosphoproteins")
plot_comparison(venn_guy11, "peptides", "Guy11 across experiments: peptides")
plot_comparison(venn_guy11, "proteins", "Guy11 across experiments: phosphoproteins")
plot_comparison(venn_mutants, "peptides", "PMK1 vs MST7: peptides")
plot_comparison(venn_mutants, "proteins", "PMK1 vs MST7: phosphoproteins")
plot_comparison(venn_mutants_guy11_shared, "peptides",
                "MST7 vs PMK1 among Guy11-shared peptides")
plot_comparison(venn_mutants_guy11_shared, "proteins",
                "MST7 vs PMK1 among Guy11-shared phosphoproteins")
```

## Save Venn diagrams as PNG files

```{r save-venns}
venn_plots <- list(
  pmk1_peptides = list(comparison = venn_pmk1, entity = "peptides",
                       title = "PMK1 experiment: peptides"),
  pmk1_phosphoproteins = list(comparison = venn_pmk1, entity = "proteins",
                              title = "PMK1 experiment: phosphoproteins"),
  mst7_peptides = list(comparison = venn_mst7, entity = "peptides",
                       title = "MST7 experiment: peptides"),
  mst7_phosphoproteins = list(comparison = venn_mst7, entity = "proteins",
                              title = "MST7 experiment: phosphoproteins"),
  guy11_peptides = list(comparison = venn_guy11, entity = "peptides",
                        title = "Guy11 across experiments: peptides"),
  guy11_phosphoproteins = list(comparison = venn_guy11, entity = "proteins",
                               title = "Guy11 across experiments: phosphoproteins"),
  mutants_peptides = list(comparison = venn_mutants, entity = "peptides",
                          title = "PMK1 vs MST7: peptides"),
  mutants_phosphoproteins = list(comparison = venn_mutants, entity = "proteins",
                                title = "PMK1 vs MST7: phosphoproteins")
  ,mutants_guy11_shared_peptides = list(
    comparison = venn_mutants_guy11_shared, entity = "peptides",
    title = "MST7 vs PMK1 among Guy11-shared peptides"
  )
  ,mutants_guy11_shared_phosphoproteins = list(
    comparison = venn_mutants_guy11_shared, entity = "proteins",
    title = "MST7 vs PMK1 among Guy11-shared phosphoproteins"
  )
)

for (plot_name in names(venn_plots)) {
  plot_spec <- venn_plots[[plot_name]]
  grDevices::png(
    filename = file.path("results", paste0("venn_", plot_name, ".png")),
    width = 7, height = 6, units = "in", res = 300
  )
  plot_comparison(plot_spec$comparison, plot_spec$entity, plot_spec$title)
  grDevices::dev.off()
}
```


## Guy11 within-experiment kinetics, then between-experiment correlation (pepdiff bootstrap-t)

Guy11 (wild type) was run as the shared control genotype in both the mst7
experiment and the pmk1 experiment -- two separate MS submissions/batches.
This asks: within each experiment, which Guy11 phosphosites change over
the timecourse relative to 0h, and for the ones that do, does the *shape*
of that response over time (its kinetics) reproduce between the two
experiments?

1. Within each experiment separately, test every timepoint (1h, 1.5h, 2h,
   4h, 6h) against 0h for Guy11, using
   [pepdiff](https://github.com/TeamMacLean/pepdiff)'s bootstrap-t test
   with 10,000 iterations. Keep phosphosites with `|log2FC| >= 1 & p <
   0.05` at any comparison, in either experiment.
2. For that significant set, correlate each phosphosite's actual
   mean-intensity trajectory across all 6 timepoints between the two
   experiments (Pearson r per peptide).

This uses VSN-normalized, **imputed** intensities, and `Modified Sequence`
as the peptide/site identifier (`Peptide Sequence` alone is not unique per
row -- see the script header for details, including a caveat on
MNAR-imputed cells). `pepdiff::compare()` hardcodes `n_boot = 1000` and
doesn't expose it, so this calls `pepdiff::test_bootstrap_t()` directly;
see `analysis/scripts/phosphoproteome_diffexp_bootstrap.r` for the full
rationale.

**This is a multi-hour computation** (~16,000 shared phosphosites x 5
timepoints x 2 experiments x 10,000 bootstrap iterations each -- every one
of them runs the full bootstrap, since imputed data has no missing values
to short-circuit on). It is not intended to be re-run live inside an
interactive knit -- run `analysis/scripts/run_guy11_bootstrap_diffexp.r`
once via `Rscript` (results are checkpointed per comparison in
`results/guy11_bootstrap_checkpoints/` so an interrupted run can resume;
a `.sbatch` version is also available for running it on the HPC instead)
and just reload the saved results here.

```{r guy11-bootstrap-diffexp, eval=FALSE}
# Equivalent to running: Rscript analysis/scripts/run_guy11_bootstrap_diffexp.r
source("analysis/scripts/run_guy11_bootstrap_diffexp.r")
```

```{r guy11-bootstrap-diffexp-reload}
guy11_timepoint_vs_0h <- readxl::read_xlsx("results/guy11_timepoint_vs_0h_bootstrap.xlsx")
guy11_kinetics_cor <- readxl::read_xlsx("results/guy11_kinetics_correlation.xlsx")
knitr::kable(summary(guy11_kinetics_cor$r))
```

### Sanity checks: volcano plots and kinetics correlation

```{r guy11-volcano-plots}
p_volcano <- plot_diffexp_volcano(guy11_timepoint_vs_0h, n_boot = 10000)
print(p_volcano)
ggplot2::ggsave("results/guy11_timepoint_vs_0h_bootstrap_volcano.png", p_volcano, width = 16, height = 7)
```

```{r guy11-kinetics-correlation-distribution}
p_kinetics_dist <- plot_kinetics_correlation_distribution(guy11_kinetics_cor)
print(p_kinetics_dist)
ggplot2::ggsave("results/guy11_kinetics_correlation_distribution.png", p_kinetics_dist, width = 10, height = 7)
```

Example trajectories: 10 randomly sampled peptides from each of three
correlation-strength buckets (low/negative r <= -0.5, no/near-zero
|r| <= 0.1, high/positive r >= 0.5) -- a visual check that the correlation
numbers reflect real agreement or disagreement in shape, not just noise.
Random sampling within each bucket (rather than the single most extreme
peptides) gives a fairer sense of what a *typical* peptide in that bucket
looks like.

```{r guy11-kinetics-examples}
guy11_long_data <- readxl::read_xlsx("results/guy11_between_experiments_long_imputed.xlsx")
kin_examples <- select_kinetics_examples(guy11_kinetics_cor, n = 10, seed = 1)

p_low <- plot_peptide_kinetics_grid(guy11_long_data, kin_examples$low, title = "10 random low/negatively correlated Guy11 peptide kinetics (mst7 vs pmk1)")
p_no <- plot_peptide_kinetics_grid(guy11_long_data, kin_examples$no, title = "10 random no/near-zero correlation Guy11 peptide kinetics")
p_high <- plot_peptide_kinetics_grid(guy11_long_data, kin_examples$high, title = "10 random high/positively correlated Guy11 peptide kinetics (mst7 vs pmk1)")

print(p_low)
print(p_no)
print(p_high)

ggplot2::ggsave("results/guy11_kinetics_examples_low.png", p_low, width = 14, height = 7)
ggplot2::ggsave("results/guy11_kinetics_examples_no.png", p_no, width = 14, height = 7)
ggplot2::ggsave("results/guy11_kinetics_examples_high.png", p_high, width = 14, height = 7)

writexl::write_xlsx(kin_examples$low, "results/guy11_kinetics_examples_low.xlsx")
writexl::write_xlsx(kin_examples$no, "results/guy11_kinetics_examples_no.xlsx")
writexl::write_xlsx(kin_examples$high, "results/guy11_kinetics_examples_high.xlsx")
```


## Notes

- The functions assume the column names contain the strings `Spectral Count` and `MaxLFQ Intensity` to detect sample columns. Adjust detection if your TSV differs.

