---
title: "Infection phosphoproteome analysis"
output: html_document
params:
  run_long_analyses: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

## Source helper scripts

```{r source-scripts}
source("analysis/scripts/phosphoproteome_cleanup.r")
source("analysis/scripts/phosphoproteome_analysis.r")
source("analysis/scripts/phosphoproteome_vsn_imputation.r")
```

## File paths

```{r paths}
pmk1_path <- file.path("raw", "pmk1", "combined_modified_peptide.tsv")
mst7_path <- file.path("raw", "mst7", "combined_modified_peptide.tsv")

# Helper to detect spectral and MaxLFQ columns from a data frame
detect_spec_and_lfq <- function(df) {
  spec_cols <- colnames(df)[stringr::str_detect(colnames(df), regex("Spectral Count", ignore_case = TRUE))]
  lfq_cols <- colnames(df)[stringr::str_detect(colnames(df), regex("MaxLFQ Intensity", ignore_case = TRUE))]
  intensity_cols <- colnames(df)[
    stringr::str_detect(colnames(df), regex(" Intensity$", ignore_case = TRUE)) &
      !stringr::str_detect(colnames(df), regex(" MaxLFQ Intensity$", ignore_case = TRUE))
  ]
  list(spec = spec_cols, lfq = lfq_cols, intensity = intensity_cols)
}
```

## Cleanup datasets

```{r cleanup}
dir.create("results", showWarnings = FALSE, recursive = TRUE)

# Load raw to detect columns
raw_pmk1 <- load_phospho_data(pmk1_path)
cols_pmk1 <- detect_spec_and_lfq(raw_pmk1)

raw_mst7 <- load_phospho_data(mst7_path)
cols_mst7 <- detect_spec_and_lfq(raw_mst7)

# Run full cleanup (phospho_all and phospho_only)
pmk1_clean <- cleanup_phospho_dataset(
  pmk1_path, mod_col = "Assigned Modifications",
  intensity_cols = cols_pmk1$intensity, min_bioreps = 2
)
mst7_clean <- cleanup_phospho_dataset(
  mst7_path, mod_col = "Assigned Modifications",
  intensity_cols = cols_mst7$intensity, min_bioreps = 2
)

# Expose datasets
pmk1_phospho_all <- pmk1_clean$phospho_all
pmk1_phospho_only <- pmk1_clean$phospho_only
pmk1_phospho_all_filtered <- pmk1_clean$phospho_all_biorep_filtered
pmk1_phospho_only_filtered <- pmk1_clean$phospho_only_biorep_filtered

mst7_phospho_all <- mst7_clean$phospho_all
mst7_phospho_only <- mst7_clean$phospho_only
mst7_phospho_all_filtered <- mst7_clean$phospho_all_biorep_filtered
mst7_phospho_only_filtered <- mst7_clean$phospho_only_biorep_filtered

writexl::write_xlsx(raw_pmk1, "results/pmk1_00_raw.xlsx")
writexl::write_xlsx(pmk1_phospho_all, "results/pmk1_01_phospho_all.xlsx")
writexl::write_xlsx(pmk1_phospho_only, "results/pmk1_02_phospho_only.xlsx")
writexl::write_xlsx(pmk1_phospho_all_filtered, "results/pmk1_03_phospho_all_min_2_bioreps.xlsx")
writexl::write_xlsx(pmk1_phospho_only_filtered, "results/pmk1_04_phospho_only_min_2_bioreps.xlsx")

writexl::write_xlsx(raw_mst7, "results/mst7_00_raw.xlsx")
writexl::write_xlsx(mst7_phospho_all, "results/mst7_01_phospho_all.xlsx")
writexl::write_xlsx(mst7_phospho_only, "results/mst7_02_phospho_only.xlsx")
writexl::write_xlsx(mst7_phospho_all_filtered, "results/mst7_03_phospho_all_min_2_bioreps.xlsx")
writexl::write_xlsx(mst7_phospho_only_filtered, "results/mst7_04_phospho_only_min_2_bioreps.xlsx")
```

## VSN normalization and imputation

```{r vsn-imputation}
pmk1_vsn <- process_vsn_imputation(
  "results/pmk1_03_phospho_all_min_2_bioreps.xlsx", "pmk1"
)
mst7_vsn <- process_vsn_imputation(
  "results/mst7_03_phospho_all_min_2_bioreps.xlsx", "mst7"
)
```

```{r load-vsn-stages}
pmk1_vsn_normalized <- readxl::read_xlsx("results/pmk1_05_vsn_normalized_nonimputed.xlsx")
pmk1_imputed <- readxl::read_xlsx("results/pmk1_05_vsn_mar_mnar_imputed.xlsx")
mst7_vsn_normalized <- readxl::read_xlsx("results/mst7_05_vsn_normalized_nonimputed.xlsx")
mst7_imputed <- readxl::read_xlsx("results/mst7_05_vsn_mar_mnar_imputed.xlsx")

pmk1_sample_cols <- stringr::str_remove(cols_pmk1$intensity, " Intensity$")
mst7_sample_cols <- stringr::str_remove(cols_mst7$intensity, " Intensity$")
```

## Principal component analysis

PCA is run separately for PMK1 and MST7 using their VSN-normalized,
imputed phosphosite intensities. Samples are the observations and
phosphosites are the variables; phosphosites are not individually scaled
because the inputs are already VSN-normalized. Incomplete or zero-variance
features are excluded. Score plots show genotype by color and timepoint by
shape; PCA scores, feature loadings, and variance explained are saved in
experiment-specific workbooks.

```{r phosphoproteome-pca}
pca_figure_dir <- file.path("results", "figures", "pca")

pmk1_pca <- run_phosphoproteome_pca(
  pmk1_imputed, pmk1_sample_cols, experiment = "pmk1",
  output_dir = pca_figure_dir,
  workbook_path = "results/pmk1_pca.xlsx"
)
mst7_pca <- run_phosphoproteome_pca(
  mst7_imputed, mst7_sample_cols, experiment = "mst7",
  output_dir = pca_figure_dir,
  workbook_path = "results/mst7_pca.xlsx"
)

pmk1_pca$score_plot
pmk1_pca$scree_plot
mst7_pca$score_plot
mst7_pca$scree_plot
```

## Summaries and plots (phospho_all)

```{r summaries-phospho-all}
# Summarise per-sample metrics for phospho_all
pmk1_summary_all <- summarise_phospho_samples(pmk1_phospho_all, spec_cols = cols_pmk1$spec, lfq_cols = cols_pmk1$lfq)
mst7_summary_all <- summarise_phospho_samples(mst7_phospho_all, spec_cols = cols_mst7$spec, lfq_cols = cols_mst7$lfq)

# Example plots
plot_total_spectral_counts(pmk1_summary_all)
plot_total_maxlfq(pmk1_summary_all)
plot_total_phospho_peptides(pmk1_summary_all)
plot_total_phospho_proteins(pmk1_summary_all)

plot_total_spectral_counts(mst7_summary_all)
plot_total_maxlfq(mst7_summary_all)
plot_total_phospho_peptides(mst7_summary_all)
plot_total_phospho_proteins(mst7_summary_all)
```

```{r save-summaries-phospho-all}
qc_figure_dir <- file.path("results", "figures", "qc")
dir.create(qc_figure_dir, recursive = TRUE, showWarnings = FALSE)

## ---- Save summaries for phospho_all ----

write.csv(pmk1_summary_all, "results/pmk1_summary_all.csv", row.names = FALSE)
write.csv(mst7_summary_all, "results/mst7_summary_all.csv", row.names = FALSE)

# PMK1 plots
ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_spectral_counts_all.png"),
                plot_total_spectral_counts(pmk1_summary_all), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_maxlfq_all.png"),
                plot_total_maxlfq(pmk1_summary_all), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_phospho_peptides_all.png"),
                plot_total_phospho_peptides(pmk1_summary_all), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_phospho_proteins_all.png"),
                plot_total_phospho_proteins(pmk1_summary_all), width = 7, height = 5)

# MST7 plots
ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_spectral_counts_all.png"),
                plot_total_spectral_counts(mst7_summary_all), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_maxlfq_all.png"),
                plot_total_maxlfq(mst7_summary_all), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_phospho_peptides_all.png"),
                plot_total_phospho_peptides(mst7_summary_all), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_phospho_proteins_all.png"),
                plot_total_phospho_proteins(mst7_summary_all), width = 7, height = 5)
```

## Summaries and plots (phospho_only)

```{r summaries-phospho-only}
# Summarise per-sample metrics for phospho_only
pmk1_summary_only <- summarise_phospho_samples(pmk1_phospho_only, spec_cols = cols_pmk1$spec, lfq_cols = cols_pmk1$lfq)
mst7_summary_only <- summarise_phospho_samples(mst7_phospho_only, spec_cols = cols_mst7$spec, lfq_cols = cols_mst7$lfq)

# Example plots
plot_total_spectral_counts(pmk1_summary_only)
plot_total_maxlfq(pmk1_summary_only)
plot_total_phospho_peptides(pmk1_summary_only)
plot_total_phospho_proteins(pmk1_summary_only)

plot_total_spectral_counts(mst7_summary_only)
plot_total_maxlfq(mst7_summary_only)
plot_total_phospho_peptides(mst7_summary_only)
plot_total_phospho_proteins(mst7_summary_only)
```

```{r save-summaries-phospho-only}
## ---- Save summaries for phospho_only ----

write.csv(pmk1_summary_only, "results/pmk1_summary_only.csv", row.names = FALSE)
write.csv(mst7_summary_only, "results/mst7_summary_only.csv", row.names = FALSE)

# PMK1 plots
ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_spectral_counts_only.png"),
                plot_total_spectral_counts(pmk1_summary_only), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_maxlfq_only.png"),
                plot_total_maxlfq(pmk1_summary_only), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_phospho_peptides_only.png"),
                plot_total_phospho_peptides(pmk1_summary_only), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "pmk1_total_phospho_proteins_only.png"),
                plot_total_phospho_proteins(pmk1_summary_only), width = 7, height = 5)

# MST7 plots
ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_spectral_counts_only.png"),
                plot_total_spectral_counts(mst7_summary_only), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_maxlfq_only.png"),
                plot_total_maxlfq(mst7_summary_only), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_phospho_peptides_only.png"),
                plot_total_phospho_peptides(mst7_summary_only), width = 7, height = 5)

ggplot2::ggsave(file.path(qc_figure_dir, "mst7_total_phospho_proteins_only.png"),
                plot_total_phospho_proteins(mst7_summary_only), width = 7, height = 5)
```

The cleaned PMK1 and MST7 tables are already saved as Excel workbooks in the
cleanup section. Duplicate CSV exports are omitted because they create very
large result files without adding analysis inputs.

## Venn analyses

```{r venn-sets}
# A feature is detected when at least one sample for that strain has a count > 0.
features_for_strain <- function(df, spec_cols, strain, id_col) {
  strain_cols <- spec_cols[stringr::str_detect(
    tolower(spec_cols), paste0("^", stringr::fixed(tolower(strain)), "_")
  )]
  if (length(strain_cols) == 0) stop("No spectral-count columns found for ", strain)

  df %>%
    filter(rowSums(across(all_of(strain_cols), ~ replace_na(as.numeric(.x), 0))) > 0) %>%
    pull(all_of(id_col)) %>%
    as.character() %>%
    na.omit() %>%
    unique()
}

make_comparison <- function(df_a, spec_a, strain_a, df_b, spec_b, strain_b,
                            label_a = strain_a, label_b = strain_b) {
  setNames(list(
    peptides = setNames(list(
      features_for_strain(df_a, spec_a, strain_a, "Peptide Sequence"),
      features_for_strain(df_b, spec_b, strain_b, "Peptide Sequence")
    ), c(label_a, label_b)),
    proteins = setNames(list(
      features_for_strain(df_a, spec_a, strain_a, "Protein ID"),
      features_for_strain(df_b, spec_b, strain_b, "Protein ID")
    ), c(label_a, label_b))
  ), c("peptides", "proteins"))
}

# Within-experiment comparisons.
venn_pmk1 <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "guy11",
                             pmk1_phospho_all, cols_pmk1$spec, "pmk1")
venn_mst7 <- make_comparison(mst7_phospho_all, cols_mst7$spec, "guy11",
                             mst7_phospho_all, cols_mst7$spec, "mst7")

# Between-experiment comparisons, using all samples for each named strain.
venn_guy11 <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "guy11",
                              mst7_phospho_all, cols_mst7$spec, "guy11",
                              "Guy11 PMK1", "Guy11 MST7")
venn_mutants <- make_comparison(pmk1_phospho_all, cols_pmk1$spec, "pmk1",
                                mst7_phospho_all, cols_mst7$spec, "mst7")

# Restrict the mutant comparison to features detected in both Guy11 datasets.
guy11_shared <- list(
  peptides = intersect(venn_guy11$peptides[[1]], venn_guy11$peptides[[2]]),
  proteins = intersect(venn_guy11$proteins[[1]], venn_guy11$proteins[[2]])
)

venn_mutants_guy11_shared <- make_comparison(
  mst7_phospho_all %>%
    filter(as.character(.data[["Peptide Sequence"]]) %in% guy11_shared$peptides),
  cols_mst7$spec, "mst7",
  pmk1_phospho_all %>%
    filter(as.character(.data[["Peptide Sequence"]]) %in% guy11_shared$peptides),
  cols_pmk1$spec, "pmk1", "MST7", "PMK1"
)

venn_mutants_guy11_shared$proteins <- setNames(list(
  features_for_strain(
    mst7_phospho_all %>%
      filter(as.character(.data[["Protein ID"]]) %in% guy11_shared$proteins),
    cols_mst7$spec, "mst7", "Protein ID"
  ),
  features_for_strain(
    pmk1_phospho_all %>%
      filter(as.character(.data[["Protein ID"]]) %in% guy11_shared$proteins),
    cols_pmk1$spec, "pmk1", "Protein ID"
  )
), c("MST7", "PMK1"))
```
```

## Plot proportional Venn diagrams

```{r plot-venns}
library(VennDiagram)
venn_figure_dir <- file.path("results", "figures", "venn")
dir.create(venn_figure_dir, recursive = TRUE, showWarnings = FALSE)

plot_comparison <- function(comparison, entity, title) {
  sets <- comparison[[entity]]
  counts <- c(length(sets[[1]]), length(sets[[2]]),
              length(intersect(sets[[1]], sets[[2]])))
  VennDiagram::draw.pairwise.venn(
    area1 = counts[[1]], area2 = counts[[2]], cross.area = counts[[3]],
    category = names(sets), scaled = TRUE, euler.d = TRUE,
    fill = c("#2F6690", "#F28E2B"), alpha = c(0.55, 0.55),
    cex = 1.1, cat.cex = 1.1, ind = FALSE
  )
  grid::grid.draw(grid::textGrob(
    title, y = grid::unit(0.98, "npc"),
    just = c("center", "top"),
    gp = grid::gpar(fontface = "bold", cex = 1.1)
  ))
  invisible(NULL)
}

plot_comparison(venn_pmk1, "peptides", "PMK1 experiment: peptides")
plot_comparison(venn_pmk1, "proteins", "PMK1 experiment: phosphoproteins")
plot_comparison(venn_mst7, "peptides", "MST7 experiment: peptides")
plot_comparison(venn_mst7, "proteins", "MST7 experiment: phosphoproteins")
plot_comparison(venn_guy11, "peptides", "Guy11 across experiments: peptides")
plot_comparison(venn_guy11, "proteins", "Guy11 across experiments: phosphoproteins")
plot_comparison(venn_mutants, "peptides", "PMK1 vs MST7: peptides")
plot_comparison(venn_mutants, "proteins", "PMK1 vs MST7: phosphoproteins")
plot_comparison(venn_mutants_guy11_shared, "peptides",
                "MST7 vs PMK1 among Guy11-shared peptides")
plot_comparison(venn_mutants_guy11_shared, "proteins",
                "MST7 vs PMK1 among Guy11-shared phosphoproteins")
```

## Save Venn diagrams as PNG files

```{r save-venns}
venn_plots <- list(
  pmk1_peptides = list(comparison = venn_pmk1, entity = "peptides",
                       title = "PMK1 experiment: peptides"),
  pmk1_phosphoproteins = list(comparison = venn_pmk1, entity = "proteins",
                              title = "PMK1 experiment: phosphoproteins"),
  mst7_peptides = list(comparison = venn_mst7, entity = "peptides",
                       title = "MST7 experiment: peptides"),
  mst7_phosphoproteins = list(comparison = venn_mst7, entity = "proteins",
                              title = "MST7 experiment: phosphoproteins"),
  guy11_peptides = list(comparison = venn_guy11, entity = "peptides",
                        title = "Guy11 across experiments: peptides"),
  guy11_phosphoproteins = list(comparison = venn_guy11, entity = "proteins",
                               title = "Guy11 across experiments: phosphoproteins"),
  mutants_peptides = list(comparison = venn_mutants, entity = "peptides",
                          title = "PMK1 vs MST7: peptides"),
  mutants_phosphoproteins = list(comparison = venn_mutants, entity = "proteins",
                                title = "PMK1 vs MST7: phosphoproteins")
  ,mutants_guy11_shared_peptides = list(
    comparison = venn_mutants_guy11_shared, entity = "peptides",
    title = "MST7 vs PMK1 among Guy11-shared peptides"
  )
  ,mutants_guy11_shared_phosphoproteins = list(
    comparison = venn_mutants_guy11_shared, entity = "proteins",
    title = "MST7 vs PMK1 among Guy11-shared phosphoproteins"
  )
)

for (plot_name in names(venn_plots)) {
  plot_spec <- venn_plots[[plot_name]]
  grDevices::png(
    filename = file.path(venn_figure_dir, paste0("venn_", plot_name, ".png")),
    width = 7, height = 6, units = "in", res = 300
  )
  plot_comparison(plot_spec$comparison, plot_spec$entity, plot_spec$title)
  grDevices::dev.off()
}
```

## Guy11 differential-expression and significant-site correlations

The 10,000-iteration bootstrap differential-expression analysis can take
several hours. Set `run_long_analyses: true` in the R Markdown parameters to
run it together with the subsequent non-imputed VSN correlations, filtered
to sites with `|log2FC| >= 1` and raw `p < 0.05` at any timepoint in either
experiment. The bootstrap script checkpoints completed comparisons to allow
resumption after interruption.

```{r long-analyses, eval=params$run_long_analyses}
source("analysis/scripts/run_guy11_bootstrap_diffexp.r")
source("analysis/scripts/run_guy11_filtered_vsn_correlation.r")

guy11_vsn_significant_peptide_cor <- readxl::read_xlsx(
  "results/guy11_vsn_filtered_significant_peptide_kinetics_correlation.xlsx"
)
guy11_vsn_significant_timepoint_cor <- readxl::read_xlsx(
  "results/guy11_vsn_filtered_significant_timepoint_pair_correlations.xlsx"
)
guy11_vsn_significant_model_cor <- readxl::read_xlsx(
  "results/guy11_vsn_filtered_significant_model_predicted_kinetics_correlation.xlsx"
)
knitr::kable(summary(guy11_vsn_significant_peptide_cor$r))
knitr::kable(summary(guy11_vsn_significant_peptide_cor$spearman_r))
knitr::kable(summary(guy11_vsn_significant_model_cor$r))
knitr::kable(guy11_vsn_significant_timepoint_cor)
```


## Notes

- The functions assume the column names contain the strings `Spectral Count` and `MaxLFQ Intensity` to detect sample columns. Adjust detection if your TSV differs.
