---
title: "Modular RNA-seq pipeline for mst7 transcriptome"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

```{r package-check, warning=TRUE}
required_pkgs <- c(
  "dplyr", "readr", "purrr", "tidyr", "stringr", "tibble",
  "here", "ggplot2", "reshape2", "sleuth", "UpSetR", "RankProd"
)
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing_pkgs) > 0) {
  warning("Note: Some packages are not installed. Pipeline will skip related sections.\n", 
          "Missing: ", paste(missing_pkgs, collapse = ", "))
}
```

# Overview

This document runs the modular RNA-seq workflow for the mst7 transcriptome project. It reads the sample metadata, sources the reusable analysis scripts, runs sleuth-based DGE comparisons from the comparison table, generates DEG summary plots, creates upset plots, and generates correlation matrices.

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(purrr)
library(tidyr)
library(stringr)
library(tibble)
library(here)
library(ggplot2)
library(reshape2)
```

```{r paths}
project_dir <- here::here()
analysis_dir <- file.path(project_dir, "analysis")
scripts_dir <- file.path(analysis_dir, "scripts")
results_dir <- file.path(project_dir, "results")
fig_dir <- file.path(results_dir, "figures")
tables_dir <- file.path(results_dir, "tables")
dge_dir <- file.path(results_dir, "dge")
upset_dir <- file.path(fig_dir, "upset")
cor_dir <- file.path(fig_dir, "correlation")

for (dir_path in c(fig_dir, tables_dir, dge_dir, upset_dir, cor_dir)) {
  dir.create(dir_path, recursive = TRUE, showWarnings = FALSE)
}
```

```{r source-scripts, message=FALSE}
safe_source <- function(file) {
  tryCatch(
    source(file, local = FALSE),
    error = function(e) {
      warning("Failed to source ", basename(file), ": ", e$message)
    }
  )
}

cat("\nSourceing utility scripts...\n")
safe_source(file.path(scripts_dir, "comparison_table.R"))
safe_source(file.path(scripts_dir, "upset_functions.R"))
safe_source(file.path(scripts_dir, "correlation_functions.R"))
safe_source(file.path(scripts_dir, "dge_functions.R"))
safe_source(file.path(scripts_dir, "rank_products_functions.R"))
cat("Scripts sourced successfully.\n")
```

```{r metadata}
s2c <- read_csv(file.path(project_dir, "raw", "run_metadata.txt"), show_col_types = FALSE) %>%
  mutate(
    path = file.path(project_dir, "raw", path),
    name = str_remove(name, "_0h_\\d+$"),
    sample = paste0(name, "_", timepoint, "_", biorep),
    name = factor(name),
    timepoint = factor(timepoint),
    biorep = factor(biorep)
  ) %>%
  mutate(
    experiment = if_else(name %in% c("pmk1", "Guy11M"), "Exp2", "Exp1"),
    experiment = factor(experiment, levels = c("Exp1", "Exp2"))
  )
```

# Sleuth differential expression

```{r run-dge, message=FALSE, warning=FALSE}
tryCatch(
  {
    cat("Running sleuth DGE comparisons...\n")
    dge_results <- run_comparisons(s2c, comparisons, out_dir = dge_dir)
    cat("DGE completed. Found", nrow(dge_results), "results.\n")
  },
  error = function(e) {
    warning("DGE analysis failed (sleuth may not be installed): ", e$message)
    dge_results <<- tibble()
  }
)
```

```{r dge-summary-plots}
if (exists('dge_results') && nrow(dge_results) > 0) {
  tryCatch(
    {
      for (lab in unique(dge_results$comparison_label)) {
        if (!is.na(lab) && nzchar(lab)) {
          plot_dge_summary(dge_results, lab, out_dir = fig_dir)
        }
      }
    },
    error = function(e) {
      warning("DGE plot generation failed: ", e$message)
    }
  )
} else {
  cat("Skipping DGE plots (no DGE results available).\n")
}
```

# Upset plots

```{r upset-plots}
tryCatch(
  {
    cat("\nGenerating upset plots...\n")
    deg_sets <- load_deg_tables(
      comparison_labels = comparisons$label,
      dge_dir = dge_dir,
      qval_threshold = 0.05,
      fc_threshold = 0
    )
    
    if (length(deg_sets) > 0) {
      plot_upset(deg_sets, label = "all_comparisons", out_dir = upset_dir)
      cat("Upset plots completed.\n")
    } else {
      cat("No DEG sets found for upset plots.\n")
    }
  },
  error = function(e) {
    warning("Upset plot generation failed: ", e$message)
  }
)
```

# Correlation matrices

```{r correlation-matrices}
tryCatch(
  {
    cat("\nGenerating correlation matrices...\n")
    tpm_matrix <- load_tpm_matrix(file.path(project_dir, "raw", "all_samples_tpm_matrix.txt"))
    
    tryCatch(cor_matrix_all(tpm_matrix, s2c), error = function(e) warning("All samples correlation failed: ", e$message))
    tryCatch(cor_matrix_exp1(tpm_matrix, s2c), error = function(e) warning("Exp1 correlation failed: ", e$message))
    tryCatch(cor_matrix_exp2(tpm_matrix, s2c), error = function(e) warning("Exp2 correlation failed: ", e$message))
    tryCatch(cor_matrix_key_strains(tpm_matrix, s2c), error = function(e) warning("Key strains correlation failed: ", e$message))
    
    cat("Correlation matrices completed.\n")
  },
  error = function(e) {
    warning("Correlation analysis failed: ", e$message)
  }
)
```

# Rank-products plots

```{r rank-products}
tryCatch(
  {
    cat("\nRunning rank products analysis...\n")
    rp_results <- run_rank_products_comparisons(s2c, comparisons, out_dir = file.path(results_dir, "rank_products"))
    
    if (nrow(rp_results) > 0) {
      cat("Generating rank products plots...\n")
      for (lab in unique(rp_results$comparison_label)) {
        if (!is.na(lab) && nzchar(lab)) {
          tryCatch(
            plot_rank_products_mirrored(
              rp_results,
              lab,
              out_dir = file.path(fig_dir, "rank_products")
            ),
            error = function(e) warning("Plot for ", lab, " failed: ", e$message)
          )
        }
      }
      cat("Rank products analysis completed.\n")
    } else {
      cat("No rank products results generated.\n")
    }
  },
  error = function(e) {
    warning("Rank products analysis failed (RankProd may not be installed): ", e$message)
  }
)
```

# Session info

```{r session-info}
sessionInfo()
```
